### 2016-01-18 papers/CAFe/HIPS-paper/coarray-extensions.tex
Near final pass.

Topics highlighted in this section are: 1. Distributed memory array allocation;
2. Explicit memory placement; 3. Remote memory transfer; and 4. Remote execution.  This
description is within the context of extensions to Fortran; as shorthand, these extensions
are referred to as CAFe, for Coarray Fortran extensions.  CAFe is complementary to
previous work extending coarray Fortran\cite{mellor-crummey:2009:caf2,jin:2011:caf2}.
Please note that throughout this paper, text appearing in all caps (such as
\texttt{GET\_SUBIMAGE}) indicates new syntax or functionality extended by CAFe.

@@

In Fortran, arrays that are visible to other program images must be declared with the
codimension attribute, for example,
%%\small
\begin{verbatim}
real, allocatable :: U(:,:)[:], V(:,:)[:]
\end{verbatim}
%%\normalsize
declares that coarrays \texttt{U} and \texttt{V} are allocatable with a rank of two and a
corank of 1.  In Fortran, square brackets \texttt{[ ]} are used to indicate operations
(possibly expensive) on distributed memory where parentheses \texttt{( )} are used to
select specific array elements; square brackets, on the other hand, denote the
\emph{location} of the array or array elements selected.

@@

As discussed, CAFe allows coarray memory to be conditionally allocated on a subimage.
It should be allocated conditionally because the requested subimage in a \texttt{GET\_SUBIMAGE}
call may not be available.  In addition, not all coarrays need be allocated on each
subimage.  For example, the following code segment allocates memory for coarrays
\texttt{U} and \texttt{V} on all images, but only coarray \texttt{U} is allocated on
the subimage specified by \texttt{device}:
%%\small
\begin{verbatim}
   allocate(U(M,N)[*], V(M,N)[*])
   if (device \= this_image()) then
     allocate(U(M,N)[*])  [[device]]
   end if
\end{verbatim}
%%\normalsize
where \texttt{M} and \texttt{N} are constant parameters.  The first allocate statement
allocates coarrays \texttt{U} and \texttt{V} on all program images.  The \texttt{*} symbol
in the allocation is required for the last corank dimension in CAF (in this case there is
only one) to allow for the number of program images to be a runtime parameter, which can
be ascertained with a \texttt{num\_images()} function call.  If \texttt{device} is
available, its numeric value will be different from \texttt{this\_image()} and thus
coarray \texttt{U} will also be allocated on the device.  Note that placement of the
memory is specified by the use of double square bracket notation \texttt{[[ ]]}.  Like
single square brackets, double square brackets indicate something special (and possibly
expensive) is to occur, in this case possibly remote memory allocation.

Deallocation of subimage memory is similar to allocation,
%%\small
\begin{verbatim}
   if (device \= this_image()) then
     deallocate(U)  [[device]]
   end if
\end{verbatim}
%%\normalsize

@@

However, CAFe allows images to execute tasks on hosted subimages using standard procedure
calls.  For example,
%%\small
\begin{verbatim}
  call relax(U[device])  [[device]]
\end{verbatim}
%%\normalsize
executes the subroutine relax, on the subimage device, using array \texttt{U} \emph{located}
on the same subimage.  This follows the same pattern as introduced for remote coarray allocation,
whereby the double square bracket notation indicates \emph{where} the function will execute.
Functions to be executed on a subimage must be pure procedures (i.e., declared with the Fortran
keyword \texttt{pure}) and scalar formal parameters must be passed by value (i.e., declared with the
\texttt{value} attribute).

@@

The comparison with OpenMP also suggests that subimage execution may not be restricted to
a single hardware execution unit, but may indeed be executed by a team of threads in a
single (shared) memory address space.  This association with a team of threads
will be exploited below when considering a team of images.

@@

The most important concept introduced by the 2015 standard is that of a team of images.
However, subimages may not be included within a team of standard CAF images.  However, if one
considers the comparison of the execution of a \texttt{do concurrent} block on a subimage
with a team of OpenMP threads (as described above), the subimage could be considered as
a team of one member that \emph{may} be executing its tasks on a team of shared memory threads.
This allows allows the use of a \texttt{sync team} statement to synchronize with
possibly executing threads.

Thus we introduce another CAFe function, \texttt{GET\_SUBIMAGE\_TEAM()}, that returns a
variable representing a fully-formed (established) team consisting of only one team
member, the current subimage.  If this function is called outside the context of a task
executing on a subimage it will return only a team consisting of the current image,
\texttt{this\_image()}.  Synchronization across possible subimage threads is then possible
using the statement
%%\small
\begin{verbatim}
  sync team (GET_SUBIMAGE_TEAM())
\end{verbatim}
%%\normalsize
This allows implementation of the technique of double buffering within a subimage task
so that all of the \emph{potential} subimage threads will have completed operating on the
temporary buffer before it is copied back to the primary array variable.

@@

In this example, the syntax \texttt{[[dev1, WITH\_EVENT=evt]]} indicates that the called
subroutine is to be run on \texttt{dev1} and that the event \texttt{evt} will receive
notification (an implicit posting) once the task has completed.  The \texttt{event wait}
statement will cause execution on the hosting image to wait until \emph{both} tasks have
completed.  One may even run multiple tasks on the \emph{same} device using events,
however unless the subimage hardware allows concurrent execution of multiple tasks, only
one task will be executed at a time.

@@

Collective operations, however, could potentially be very useful and fit nicely within
CAFe concepts.  For example, a team consisting of hosted subimages could be formed and
then data broadcasted to team members and reduction operations such as \texttt{co\_add}
performed.  However since all team members of the current team must participate in
collective operations \emph{and} since CAFe restricts subimages from being part of the
same team with regular images, it makes it impossible for an image to participate with a
team of its hosted subimages in collective operations.  So for now, collectives are
excluded from CAFe.

>>>>

Topics highlighted in this section are: 1. Distributed memory array allocation;
2. Explicit memory placement; 3. Remote memory transfer; and 4. Remote execution
and synchronization.  This
description is within the context of extensions to Fortran; as shorthand, these extensions
are referred to as CAFe, for Coarray Fortran extensions.  CAFe is complementary to
previous work extending coarray Fortran\cite{mellor-crummey:2009:caf2,jin:2011:caf2}.
Please note that throughout this paper, text appearing in all caps (such as
\texttt{GET\_SUBIMAGE}) indicates new syntax or functionality extended by CAFe.

@@

In Fortran, arrays that are visible to other program images must be declared with the
codimension attribute, for example,
%%\small
\begin{verbatim}
real, allocatable :: U(:,:)[:], V(:,:)[:]
\end{verbatim}
%%\normalsize
declares that coarrays \texttt{U} and \texttt{V} are allocatable with a rank of two and a
corank of one.  In Fortran, square brackets \texttt{[ ]} are used to indicate operations
(possibly expensive) on distributed memory where parentheses \texttt{( )} are used to
select specific array elements; square brackets, on the other hand, denote the
\emph{location} of the array or array elements selected.

@@

As discussed, CAFe allows coarray memory to be conditionally allocated on a subimage.
It should be allocated conditionally because the requested subimage in a \texttt{GET\_SUBIMAGE}
call may not be available.  In addition, not all coarrays need be allocated on each
subimage.  For example, the following code segment allocates memory for coarrays
\texttt{U} and \texttt{V} on all images, but only coarray \texttt{U} is allocated on
the subimage specified by the variable \texttt{device}:
%%\small
\begin{verbatim}
   allocate(U(M,N)[*], V(M,N)[*])
   if (device /= this_image()) then
     allocate(U(M,N)[*])  [[device]]
   end if
\end{verbatim}
%%\normalsize
where \texttt{M} and \texttt{N} are constant parameters.  The first allocate statement
allocates coarrays \texttt{U} and \texttt{V} on all program images.  The \texttt{*} symbol
in the allocation is required for the last corank dimension in CAF (in this case there is
only one) to allow for the number of program images to be a runtime parameter, which can
be ascertained with a \texttt{num\_images()} function call.  If \texttt{device} is
available, its numeric value will be different from \texttt{this\_image()} and thus
coarray \texttt{U} will also be allocated on the device.  Note that placement of the
memory is specified by the use of double square bracket notation \texttt{[[ ]]}.  Like
single square brackets, double square brackets indicate something special (and possibly
expensive) is to occur, in this case possibly remote memory allocation.

Deallocation of subimage memory is similar to allocation,
%%\small
\begin{verbatim}
   if (device /= this_image()) then
     deallocate(U)  [[device]]
   end if
\end{verbatim}
%%\normalsize

@@

However, CAFe allows images to execute tasks on hosted subimages using standard procedure
calls.  For example,
%%\small
\begin{verbatim}
  call relax(U[device])  [[device]]
\end{verbatim}
%%\normalsize
executes the subroutine relax, on the subimage \texttt{device}, using coarray \texttt{U} \emph{located}
on the same subimage.  This follows the same pattern as introduced for remote coarray allocation,
whereby the double square bracket notation indicates \emph{where} the function will execute.
Functions to be executed on a subimage must be pure procedures (i.e., declared with the Fortran
keyword \texttt{pure}) and scalar formal parameters must be passed by value (i.e., declared with the
\texttt{value} attribute).

@@

The comparison with OpenMP also suggests that subimage execution may not be restricted to
a single hardware execution unit, but may indeed be executed by a team of threads in a
single (shared) memory address space.  This association with a team of threads will be
exploited below when considering a team of subimages.

@@

The most important concept introduced by the 2015 standard is that of a team of images.
Subimages may not be included within a team of standard CAF images.  However, if one
considers the comparison of the execution of a \texttt{do concurrent} block on a subimage
with a team of OpenMP threads (as described above), the subimage could be considered as
a team of one member that \emph{may} be executing its tasks on a team of shared memory threads.
This allows the use of a \texttt{sync team} statement to synchronize with
possibly executing threads.

Thus we introduce another CAFe intrinsic function, \texttt{GET\_SUBIMAGE\_TEAM()}, that returns a
variable representing a fully-formed (established) team consisting of only one team
member, the current subimage.  If this function is called outside the context of a task
executing on a subimage it will return only a team consisting of the current image,
\texttt{this\_image()}.  Synchronization across possible subimage threads is then possible
using the statement
%%\small
\begin{verbatim}
  sync team (GET_SUBIMAGE_TEAM())
\end{verbatim}
%%\normalsize
This allows implementation of the technique of double buffering within a subimage task
so that all of the \emph{potential} subimage threads will have completed operating on the
temporary buffer before it is copied back to the primary array variable.

@@

In this example, the syntax \texttt{[[dev1, WITH\_EVENT=evt]]} indicates that the called
subroutine is to be run on \texttt{dev1} and that the event variable \texttt{evt} will receive
notification (an implicit posting) once the task has completed.  The \texttt{event wait}
statement will cause execution on the hosting image to wait until \emph{both} tasks have
completed.  One may even run multiple tasks on the \emph{same} device using events,
however unless the subimage hardware allows concurrent execution of multiple tasks, only
one task will be executed at a time.

@@

Collective operations, however, could potentially be very useful and fit nicely within
CAFe concepts.  For example, a team consisting of hosted subimages could be formed and
then data be broadcasted to team members and reduction operations such as \texttt{co\_add}
performed.  However since all team members of the current team must participate in
collective operations \emph{and} since CAFe restricts subimages from being part of the
same team with regular images, it makes it impossible for an image to participate with a
team of its hosted subimages in collective operations.  So for now, collectives are
excluded from CAFe.
